%% MethodsX methods article -- Elsevier elsarticle template
%% Structure follows the official "MethodsX methods article template
%% Version 7 (September 2024)". Section headings match the template and must
%% not be changed. All sections are mandatory unless marked OPTIONAL.
%%
%% Compile with pdfLaTeX (run twice to resolve \cite references).
%% Overleaf: New Project -> Upload Project -> Compiler = pdfLaTeX.
%%
%% LAYOUT: 'final,3p' = journal single-column look (for reading / advisor
%%   review). When you SUBMIT, change 'final' to 'review' for the
%%   double-spaced, line-numbered manuscript format Elsevier expects.
%%
%% TODO before submitting:
%%   - fill every [bracketed] placeholder (affiliations, ethics,
%%     funding, CRediT roles);
%%   - prepare the Graphical abstract as a SEPARATE image file;
%%   - delete this comment header.

\documentclass[final,3p,times]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\modulolinenumbers[5]

\journal{MethodsX}

\begin{document}

%% =====================================================================
%% ARTICLE INFORMATION
%% Title (max. 20 words), authors (mark corresponding with *), full
%% affiliations, corresponding e-mail (+ X handle), keywords (>= 3).
%% =====================================================================
\begin{frontmatter}

\title{A transfer-learning and automatic false-positive cleanup pipeline
for detecting primate vocalizations in long tropical-forest recordings}

\author[a]{Moshi Fu\corref{cor1}}
\ead{mfu39@wisc.edu}
\cortext[cor1]{Corresponding author.}

\author[a]{[Co-author Name]}

\affiliation[a]{organization={[Department], University of Wisconsin--Madison},
            addressline={[Street address]},
            city={Madison},
            postcode={[ZIP]},
            state={WI},
            country={USA}}

\affiliation[b]{organization={[Second institution, full address]},
            country={[Country]}}

%% ABSTRACT -- max. 200 words, ending with 1-3 bullet points that
%% outline the method (no technical steps).
\begin{abstract}
Passive acoustic monitoring (PAM) produces thousands of hours of
tropical-forest audio that are infeasible to annotate by hand. We describe an
end-to-end pipeline that detects primate vocalizations --- the putty-nosed
monkey (\textit{Cercopithecus nictitans}) and \textit{Colobus guereza} ---
in long field recordings from Makokou, Gabon. A four-class model is built by
pooling all putty-nosed call types into a single Cernic class at the training
level (so no softmax summation is needed at detection time) and by adding a
dedicated hard-negative ``confuser'' class for a forest sound that the detector
repeatedly mistakes for \textit{Colobus}. Two-second audio windows --- with
short clips embedded in real background noise and longer clips trimmed to their
highest-energy window --- are converted to mel-spectrogram images and
classified by a temporal-frequency CRNN head attached to a VGG19 backbone: the
feature map is split into four frequency bands, processed by per-band and
cross-band convolutions, and fed through a bidirectional LSTM to capture
spectro-temporal structure. Two-stage training (frozen base then partial
fine-tuning) on human-verified, label-audited clips reaches 96\% validation
accuracy with near-zero confusion between the two primate classes. At detection
time the confuser class is folded into the background group so it never
produces a detection, and a complementary low-frequency spectral-energy gate
suppresses high-frequency false positives. An automatic three-filter cleanup
additionally flags suspicious detections, and confirmed false positives are
recycled as hard negatives for iterative retraining.

\begin{itemize}
\item A temporal-frequency CRNN head on a VGG19 backbone for primate call
detection with two-stage transfer learning.
\item A dedicated hard-negative ``confuser'' class and a complementary
low-frequency spectral-energy gate that together suppress \textit{Colobus}
false positives without manual listening.
\item An automatic three-filter cleanup pipeline that feeds confirmed false
positives back as hard negatives for iterative retraining.
\end{itemize}
\end{abstract}

%% KEYWORDS -- minimum three, separated by \sep.
\begin{keyword}
Bioacoustics \sep Passive acoustic monitoring \sep Transfer learning \sep
Convolutional recurrent neural network \sep Mel-spectrogram \sep
False-positive filtering \sep Out-of-distribution detection \sep
\textit{Cercopithecus nictitans} \sep \textit{Colobus guereza}
\end{keyword}

\end{frontmatter}

%% =====================================================================
\section*{Related research article}
None.

%% =====================================================================
\section*{Graphical abstract}
\textit{[Submitted as a separate image file (min.\ 531 $\times$ 1328 px,
h $\times$ w; readable at 5 $\times$ 13 cm; TIFF/EPS/PDF/MS Office). Suggested
content: the pipeline --- reference clips $\rightarrow$
mel-spectrogram + temporal-frequency CRNN classifier $\rightarrow$
sliding-window detection $\rightarrow$ three-filter cleanup $\rightarrow$
hard-negative retraining loop.]}

%% =====================================================================
\section*{Specifications table}

\noindent
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\linewidth}{@{}lX@{}}
\toprule
Subject area & Agricultural and Biological Sciences \\
More specific subject area & Bioacoustics; passive acoustic monitoring;
conservation biology \\
Name of your method & VGG19 temporal-frequency CRNN with two-stage
fine-tuning, a dedicated hard-negative confuser class, a low-frequency
spectral-energy gate, and three-filter automatic false-positive cleanup for
primate vocalizations \\
Name and reference of original method & Transfer learning with the VGG19
architecture \cite{simonyan2015}; out-of-distribution detection via the
Mahalanobis distance \cite{lee2018}; audio event tagging with YAMNet trained
on AudioSet \cite{gemmeke2017,hershey2017} \\
Resource availability & Code:
\url{https://github.com/mo119m/primates-sound-detection} --- Python 3,
TensorFlow 2, librosa, scikit-learn, OpenCV, soundfile; the YAMNet filter
additionally needs tensorflow-hub and resampy. [Add a data DOI/repository
here if you deposit the reference clips.] \\
\bottomrule
\end{tabularx}

%% =====================================================================
%% BACKGROUND -- motivation and context. Max. 500 words.
%% =====================================================================
\section*{Background}

Passive acoustic monitoring is now a standard tool for surveying vocal animals
across large areas and long time spans, but it generates volumes of audio that
cannot realistically be reviewed by ear \cite{cauzinille2024}. Deep convolutional
networks applied to spectrogram ``images'' have become the dominant approach
for automatically detecting species-specific calls, and transfer learning from
large image or audio corpora lets such detectors be trained from comparatively
few annotated examples \cite{sun2022}. These approaches work well
on clean, isolated reference clips.

The difficulty in practice is precision rather than recall. A model trained on
curated clips of a target call will reliably fire on those calls, but when it
is run over continuous field recordings it also fires on the many non-target
sounds that dominate a rainforest soundscape --- insects, birds, wind, rain,
and the calls of non-target animals --- none of which are well represented in
the original training set. The result is a detector with a high false-positive
rate, and verifying every detection by hand reintroduces exactly the
manual-listening bottleneck that automation was meant to remove.

This method addresses that problem for two primate species recorded at
Makokou, Gabon: the putty-nosed monkey (\textit{Cercopithecus nictitans}),
whose repertoire includes acoustically distinct ``hack'', ``kek'', and
``pyow'' call types, and \textit{Colobus guereza}. Rather
than only training a classifier, we wrap it in (i) a detection stage with
domain-appropriate post-processing, (ii) an automatic three-filter cleanup
that flags likely false positives using signals independent of the detector
itself, and (iii) an iterative hard-negative mining loop that recycles
confirmed false positives into the background class and retrains. The
motivation is to obtain trustworthy detections from long recordings while
keeping the amount of manual listening close to zero, and to do so with a
workflow that other researchers can re-run on their own species by supplying
new reference clips and editing a single configuration file.

%% =====================================================================
%% METHOD DETAILS -- enough detail to reproduce the method. No word limit.
%% =====================================================================
\section*{Method details}

The pipeline has five stages --- preprocessing, training, detection, automatic
cleanup, and iterative retraining --- augmented by two targeted
false-positive controls for \textit{Colobus}: a dedicated hard-negative
``confuser'' class learned during training, and a low-frequency spectral-energy
gate applied as detection post-processing. All paths, species definitions, and
hyperparameters live in a single \texttt{config.py} so the workflow can be
re-targeted to new species or sites without touching the code.

\subsection*{1. Study area, recordings, and reference clips}

Field recordings are continuous WAV files collected at fixed acoustic stations
(denoted IPA1ST, IPA2ST, \dots; one subfolder per recording date) in
rainforest near Makokou, Gabon. Because the target species are diurnal and
most vocally active in the morning, only recordings whose start time falls
within a configurable window (default 05:30--10:30 local time, parsed from the
timestamp in each filename) are processed; this removes large stretches of
audio in which the species are unlikely to call and reduces both compute and
false positives.

The model is trained on short reference clips organized into four classes:
\begin{itemize}
\item \textbf{Cernic} --- all putty-nosed monkey call types (putty-nose, hack,
kek, pyow) pooled into a single class. Five source folders (one per call type
plus a \texttt{field\_confirmed} folder of recovered field calls) are mapped to
the single ``Cernic'' label via \texttt{config.SPECIES\_FOLDERS}, so that
merging happens at training time and the model directly outputs a unified
putty-nosed decision (putty-nose 25, hack 101, kek 124, pyow 101,
field-confirmed 19; 370 clips total);
\item \textbf{Colobus\_guereza} --- \textit{Colobus guereza} calls (617 clips);
\item \textbf{Colobus\_confuser} --- a dedicated hard-negative class containing
the forest sounds that earlier model versions repeatedly mis-fired as
\textit{Colobus}, mined from field detections across the development stations
and human-reviewed so that no genuine target call is included (654 clips).
Giving the confuser its own softmax output forces the model to learn the
\textit{Colobus}-vs-confuser boundary explicitly instead of diluting these few
hundred hard negatives inside the much larger generic Background class; at
detection time the class is folded back into the background group (see
the detection-grouping and hard-negative-mining stages below) so it never
produces a detection;
\item \textbf{Background} --- a pooled negative class assembled from ambient
forest noise, calls of non-target animals present at the site (e.g.\
\textit{Cercocebus torquatus}, \textit{Pan troglodytes}), and previously
misclassified clips recycled through the hard-negative mining loop
(approximately 5{,}900 clips total).
\end{itemize}

\noindent All training clips underwent human verification. Because the
Background class is loaded as a single pooled negative --- with no internal
sub-labels --- any genuine target call accidentally present in a false-positive
folder would teach the model to suppress that call. We therefore conducted a
manual label-quality audit of the auto-flagged false-positive pool: clips
confirmed by ear to be real putty-nosed calls were removed from the negative
pool (to prevent Cernic contamination) and, when unambiguous, recovered into
the dedicated \texttt{field\_confirmed} Cernic folder; the \textit{Colobus}
false positives were moved out of the generic pool into the new
\texttt{Colobus\_confuser} class. Content-duplicate clips --- the same field
detection stored more than once by different mining passes --- were removed by
hashing before training.

\subsection*{2. Audio preprocessing and mel-spectrogram representation}

All audio is resampled to 44.1\,kHz and standardized to a fixed 2.0\,s
duration. Clips shorter than the target (e.g.\ hack, kek, and pyow calls,
typically $\sim$0.3\,s) are embedded at a random position within a 2\,s
segment of real background noise randomly selected from the Background pool;
this eliminates the silence shortcut that zero-padding would introduce and
forces the model to distinguish the call from genuine ambient sound. Clips
longer than 2\,s (e.g.\ putty-nosed 5\,s field recordings) are trimmed to the
deterministic 2\,s window with the highest RMS energy
(\texttt{find\_loudest\_window}), ensuring that the segment reviewed by a
human annotator is identical to the segment used for training. Each 2\,s
waveform is converted to a mel-spectrogram with librosa \cite{mcfee2015} using
a 2048-sample FFT, 512-sample hop, 128 mel bands, and a 20--8000\,Hz analysis
range, and is expressed in decibels relative to its per-clip maximum. The
spectrogram is min--max normalized to [0, 255], resized to
224\,$\times$\,224 with bilinear interpolation, replicated across three
channels to form an RGB image, and finally scaled to [0, 1] for input to the
network. Using the same conversion at training and detection time keeps the
two feature distributions comparable.

\subsection*{3. Data augmentation}

To expand the reference set and make the classifier robust to field
conditions, each target-class spectrogram is expanded into seven training
examples (a 7$\times$ multiplier): one unaltered copy; three copies mixed with
randomly chosen background spectrograms at a signal-to-noise ratio drawn
uniformly from $-5$ to 10\,dB; one time-axis crop and one frequency-axis crop
(each removing 10--30\,\% from a randomly chosen edge, then resized back to the
original shape); and one frequency translation of $\pm$20 mel bins.
Background-class clips are not augmented, so that the negative class reflects
the true diversity of ambient sound rather than synthetic variants. The
augmentation operations are spectrogram-domain analogues of standard
standard spectrogram-domain masking and mixing techniques.

\subsection*{4. Model architecture}

The classifier couples a VGG19 convolutional backbone \cite{simonyan2015}
pretrained on ImageNet \cite{deng2009} with a custom temporal-frequency CRNN
head designed to exploit both the spectral location and the temporal rhythm of
primate calls. The motivation is that Colobus roars concentrate energy around
$\sim$512\,Hz, putty-nosed calls around $\sim$1\,kHz, and common confusers
(birds, insects) above $\sim$2\,kHz; encoding \emph{where} in frequency the
energy sits and \emph{when} it occurs lets the model discriminate species that
a global pooling layer would conflate.

The feature map from \texttt{block4\_conv4} of VGG19 (spatial size
28\,$\times$\,28, 512 channels) is extracted before the final VGG19 block. The
frequency axis (height) is split into four equal bands (each 7 bins), so that
each band covers a contiguous portion of the mel-frequency range. Within each
band, the frequency dimension is collapsed (average pooling over the 7 bins),
yielding a time-$\times$-channel sequence of shape 28\,$\times$\,512. Each
band is then processed by its own one-dimensional convolutional stream ---
Conv1D(128 filters, kernel size 3, same padding) followed by batch
normalization and ReLU --- producing four sequences of shape
28\,$\times$\,128. These are concatenated along the channel axis
(28\,$\times$\,512) and passed through a cross-band Conv1D(256, kernel 3,
same padding) + batch normalization + ReLU layer that allows the network to
learn interactions between frequency bands.

The resulting 28\,$\times$\,256 sequence is fed into a bidirectional LSTM with
128 units per direction (return sequences enabled, recurrent dropout 0.3),
which captures temporal dependencies such as the rhythmic hack--hack or
pyow--hack patterns characteristic of putty-nosed calling bouts. The LSTM
output sequence is summarized by concatenating global max pooling and global
average pooling over the time axis, producing a 512-dimensional vector that
retains both peak activation and average energy. This vector passes through
two dense layers (512 units, ReLU, dropout 0.5; then 256 units, ReLU, dropout
0.5) and a final four-way softmax output (Cernic, Colobus\_guereza,
Colobus\_confuser, Background).

\subsection*{5. Model training}

Training proceeds in two stages, following a standard transfer-learning
protocol of head training followed by partial fine-tuning.

\paragraph{Stage 1 --- Head training.}
The entire VGG19 convolutional base is frozen and only the temporal-frequency
CRNN head is trained. The optimizer is Adam with an initial learning rate of
$1 \times 10^{-4}$, and the loss is sparse categorical cross-entropy.

\paragraph{Stage 2 --- Partial fine-tuning.}
The last two convolutional blocks of VGG19 (\texttt{block4} and
\texttt{block5}) are unfrozen while earlier blocks remain frozen to preserve
low-level features. The learning rate is reduced to $1 \times 10^{-5}$ to
avoid catastrophic forgetting.

\paragraph{Shared training settings.}
Both stages use the augmented dataset split 80/20 into training and validation
sets with stratification by class and a fixed random seed for
reproducibility. Balanced class weights (inversely proportional to class
frequency) compensate for the larger Background class. Training runs for up to
50 epochs per stage with a batch size of 32, governed by three callbacks:
early stopping on validation accuracy (patience 10, best weights restored),
model checkpointing on best validation accuracy, and learning-rate reduction
on plateau (factor 0.5, patience 5, floor $1 \times 10^{-7}$). The best
checkpoint is saved as \texttt{best\_model.h5}. Training concludes with a
per-class accuracy report and a confusion matrix on the validation set.

\subsection*{6. Sliding-window detection}

A trained model is applied to each long recording with a 2.0\,s window
advanced in 1.0\,s steps (50\,\% overlap); a final window is anchored at the
end of the file so trailing audio is not dropped. Every window is converted to
a mel-spectrogram image and classified. The four softmax classes are first
collapsed onto detection \emph{groups} through \texttt{DETECTION\_GROUPS}: the
putty-nosed call types are already merged into Cernic at training time, so
Cernic and Colobus\_guereza map to themselves, while the Colobus\_confuser
class is routed into the Background group. Summing the confuser probability
into Background means a window the model calls ``confuser'' is excluded from
detections exactly like Background, and --- equally important --- the confuser
probability is kept \emph{out} of the Colobus\_guereza group, so a genuine
\textit{Colobus} window is no longer inflated by confuser energy. The window is
assigned to its highest-scoring group and is kept as a detection if that group
is not Background and its confidence meets a threshold (default 0.4).
Overlapping detections of the same group are then reduced by per-class
non-maximum suppression at an intersection-over-union threshold of 0.5.
Per-file detections (start time, end time, species, confidence) are written to
CSV, and the 2\,s waveform of every detection is exported as a clip so that
review and downstream filtering do not require re-reading the long recordings.

\paragraph{Low-frequency spectral-energy gate.}
\textit{Colobus guereza} roars concentrate their energy at low frequency,
whereas the residual \textit{Colobus} false positives are dominated by
high-frequency tonal bands (insect and bird sounds well above 1.5\,kHz). We
exploit this with a post-processing gate applied only to Colobus detections:
for each detection clip we compute the short-time Fourier transform (the same
2048-sample FFT and 512-sample hop used elsewhere) and take the fraction of
total in-band (20--8000\,Hz) energy that falls below a 1.5\,kHz cutoff. A
detection is retained only if this low-frequency energy ratio is at least 0.40,
and is otherwise discarded as a false positive. The threshold was set to the
5th percentile of the low-frequency energy ratio measured on the curated
\textit{Colobus} reference clips, so the gate retains roughly 95\,\% of genuine
\textit{Colobus} calls while removing the high-frequency confusers that the
softmax classifier still mislabels. Because the gate operates on the exported
detection clips, it requires no model retraining and no re-processing of the
long recordings, and its cutoff and threshold can be re-tuned at any time.

\subsection*{7. Automatic false-positive cleanup}

The detector's raw output contains many false positives from sounds absent
from the training set. Three independent filters, each exploiting a different
signal, are run over the saved detections; their agreement determines how a
detection is treated. For each detection a 2\,s clip is extracted from the
source recording.

\paragraph{(a) Mahalanobis out-of-distribution filter}
Using a penultimate dense layer as a feature extractor, per-class feature means
and a pooled, ridge-regularized inverse covariance are estimated from the
original (non-augmented) training clips \cite{lee2018}. For each detection the
Mahalanobis distance to its candidate class cluster is computed. A real call
lies close to its training cluster, whereas an unfamiliar sound does not.
Because field clips are systematically noisier than clean training clips,
thresholds calibrated on training data flag almost everything; we therefore
calibrate the cutoff on the \textbf{detection distances themselves}, flagging,
per species, the detections beyond the 95th percentile of the observed
distance distribution (a configurable percentile). This makes the filter a
relative-outlier detector that is robust to the train-to-field domain shift.

\paragraph{(b) YAMNet cross-check}
Each clip is resampled to 16\,kHz and passed to YAMNet, a network trained on
Google's AudioSet ontology of 521 everyday sound classes
\cite{gemmeke2017,hershey2017}. A detection is flagged when YAMNet's top class
is one of a curated set of non-primate sounds (birds, insects, frogs, wind,
rain, water, mechanical and human sounds, etc.); generic classes consistent
with a primate call (e.g.\ ``Animal'', ``Wild animals'') are deliberately left
unflagged. Because AudioSet has no monkey-specific class, this filter is used
only to \emph{reject} clips that are confidently something else, not to
confirm primates.

\paragraph{(c) Temporal-isolation filter}
Primates tend to call in bouts, so a genuine call usually has same-species
neighbours nearby in time. A detection with no other detection of the same
species within $\pm$30\,s (configurable) is flagged as temporally isolated.

A detection is considered \textbf{clean} only when no filter flags it,
\textbf{suspicious} when at least one filter flags it, and a \textbf{strong
false positive} when at least two filters agree. Clean and suspicious
detections are written to separate CSVs (the suspicious file carries a
human-readable \texttt{flag\_reason} column), and strong false positives are
exported as labelled WAV clips for the next stage.

\subsection*{8. Iterative hard-negative mining}

The exported strong false positives are the sounds the detector most needs to
learn to ignore. They are added to the Background class and the model is
retrained from the preprocessing stage, so that on the next pass the network
is far less likely to fire on those sounds. Repeating the detect $\rightarrow$
clean $\rightarrow$ fold-in $\rightarrow$ retrain loop a few times (typically
three to five iterations) progressively drives down the false-positive rate
without ever requiring exhaustive manual annotation of the field recordings.

When one species' false positives are dominated by a single, recurring
confuser rather than a diffuse mix of sounds, folding them into the generic
Background class is not enough: a few hundred hard negatives are easily diluted
among several thousand background clips. In that case we promote the confuser
to its own softmax class (here, \texttt{Colobus\_confuser}) and route it back
into the Background group only at detection time. This gives the optimizer an
explicit boundary to fit while leaving the detection behaviour unchanged, and
is the approach used for the low-frequency \textit{Colobus} confuser in the
current model. Because every field detection clip is exported during detection,
this hard-negative mining can be repeated from the saved clips even after the
source recordings have been deleted to save storage.

%% =====================================================================
%% METHOD VALIDATION -- data that validate the method (no Results/Discussion).
%% =====================================================================
\section*{Method validation}

We report validation at two levels: (i) performance on a held-out validation
split drawn from the same curated clip pool as training, and (ii) deployment on
the continuous field recordings, where the biology of the site provides an
independent check on the \textit{Colobus} false-positive rate.

\paragraph{Validation-set performance (V10 model).}
The four-class V10 model was evaluated on a stratified validation split of 3471
clips. Table~\ref{tab:val} summarizes per-class results and
Table~\ref{tab:confusion} gives the full confusion matrix.

\begin{table}[h]
\centering
\caption{Per-class validation-set performance of the final (V10) four-class
model.}
\label{tab:val}
\begin{tabular}{lrrr}
\toprule
Class & Samples & Accuracy (\%) & Mean confidence \\
\midrule
Cernic            & 518  & 91.89 & 0.914 \\
Colobus\_guereza  & 864  & 97.80 & 0.973 \\
Colobus\_confuser & 916  & 96.51 & 0.957 \\
Background        & 1173 & 96.50 & 0.957 \\
\midrule
\textbf{Overall}  & \textbf{3471} & \textbf{96.14} & --- \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Validation-set confusion matrix of the V10 model. Rows are true
classes, columns are predicted classes (Cer = Cernic, C\_g =
Colobus\_guereza, C\_c = Colobus\_confuser, Bg = Background).}
\label{tab:confusion}
\begin{tabular}{lrrrr}
\toprule
True $\backslash$ Pred & Cer & C\_g & C\_c & Bg \\
\midrule
Cernic            & 476 & 2   & 20  & 20   \\
Colobus\_guereza  & 2   & 845 & 5   & 12   \\
Colobus\_confuser & 11  & 3   & 884 & 18   \\
Background        & 9   & 4   & 28  & 1132 \\
\bottomrule
\end{tabular}
\end{table}

\noindent Overall accuracy is 96.14\% with a loss of 0.147. Two features of the
confusion matrix are central to the method. First, cross-species confusion
between the two primate classes is near zero: only 2 Cernic clips are called
\textit{Colobus} and 2 \textit{Colobus} clips are called Cernic (4 of 1382
primate clips). Second, the dedicated confuser class is well separated from
real \textit{Colobus}: of 864 genuine \textit{Colobus} clips only 5 are
assigned to the confuser, and of 916 confuser clips only 3 are assigned to real
\textit{Colobus} (8 of 1780, $<$0.5\,\%). Because the confuser group is folded
into Background at detection time, these confuser windows do not produce
\textit{Colobus} detections. Cernic is the most conservative class
(recall 91.9\,\%): its errors are split evenly between the confuser (20) and
Background (20) and consist mostly of weak or brief calls embedded in noise,
trading recall for precision.

\paragraph{Field deployment.}
The V10 model is applied to the continuous recordings from every available
acoustic station (16 of the 20 IPA stations are present in the archive; IPA3,
IPA5, IPA9, and IPA12 are absent). One 2\,s clip is exported for every
detection, and the low-frequency spectral-energy gate is applied to all
\textit{Colobus} detections as described above. Manual review of the
\textit{Colobus} detections confirms that they are false positives --- no
genuine \textit{C.\ guereza} vocalization was found in the analyzed recordings
--- so the field data provide a direct, fully labelled measure of
\textit{Colobus} false-positive suppression by the confuser class and the
low-frequency gate. Aggregate per-station detection counts, the Cernic
detections retained after manual review, and the number of \textit{Colobus}
false positives removed by the low-frequency gate will be reported once all
stations have been processed.

\paragraph{Architecture progression (development history).}
The components of the model were added in response to failures observed on a
development station (IPA19ST) during earlier, three-class iterations. A
frequency-band pooling head (V6) fired on essentially every loud non-target
sound; adding a temporal CRNN head with Conv1D and a bidirectional LSTM (V7)
sharply cut false positives but also suppressed real calls; splitting the
frequency axis into bands before temporal modelling (the temporal-frequency
head, V8) restored Cernic recall while keeping false positives low. The
remaining problem was specific to \textit{Colobus}: a recurring low-frequency
forest sound was repeatedly mis-fired as \textit{Colobus} and was not
controlled by folding examples into the generic Background class. Promoting that
sound to a dedicated \texttt{Colobus\_confuser} class and adding the
low-frequency gate (V10) addresses this directly, and the V10 model is the
version deployed across all stations.

%% =====================================================================
\section*{Limitations}

\begin{itemize}
\item The training set contains no real Colobus field recordings from the
Makokou site; all Colobus reference clips are sourced from curated databases,
and manual review found no genuine \textit{C.\ guereza} call in the analyzed
recordings. This domain mismatch, together with the open-set nature of the
classifier (below), drives the \textit{Colobus} false positives that the
confuser class and the low-frequency gate are designed to suppress. Collecting
site-specific Colobus recordings remains a priority for future work.
\item The classifier is closed-set: every 2\,s window is assigned to one of the
four trained classes, so a field sound unlike anything seen in training is
forced into its nearest class and can receive a high softmax score even when it
is not a genuine call. This is the root cause of the residual \textit{Colobus}
false positives, and it is why a high confidence threshold alone does not
remove them; the dedicated confuser class and the low-frequency gate target
this failure mode but cannot be expected to eliminate every out-of-distribution
sound.
\item The low-frequency gate assumes the target species' calls are
concentrated at low frequency (true for \textit{Colobus guereza} roars). It is
applied only to \textit{Colobus} detections and would need a different
frequency criterion --- or none --- for species whose calls are not
low-frequency.
\item The auto-cleanup pipeline introduces a circular bias: hard negatives
are scored by the same model that produced them, so genuine calls with low
model confidence may be systematically hidden. This risk is mitigated by a
mining-confidence audit in which high-confidence clips in the false-positive
pool are manually reviewed, and confirmed calls are recovered into the
training set.
\item The model is conservative on weak or short calls. On the validation
set, Cernic recall is 91.9\,\%; the misclassified Cernic clips --- mostly brief
hacks and pyows embedded in loud ambient noise --- are split between the
confuser class and Background. This trade-off favours precision over recall.
\item Re-targeting to new species requires curated reference clips and edits
to the configuration; performance on rare or quiet calls with few reference
examples is not guaranteed.
\item The YAMNet cross-check has no primate-specific class and only rejects
clips it can confidently attribute to a non-primate sound; tropical-forest
sounds outside AudioSet's training distribution may be mislabelled, so the
filter is conservative by design.
\item The Mahalanobis filter assumes that the majority of detections for a
species are genuine, so that outliers are informative; if a species'
detections are overwhelmingly false positives, the relative-outlier
calibration is less effective.
\item The temporal-isolation filter assumes bout-calling; genuinely isolated
or rare calls can be flagged, and very dense calling can mask true isolation.
\item A fixed 2\,s analysis window may truncate or split long vocalizations
(e.g.\ \textit{Colobus} roars), and the time-of-day filter (05:30--10:30) is
site- and species-specific and must be adjusted elsewhere.
\item The method yields detections, not calibrated abundance or density
estimates; converting counts to ecological metrics requires additional
assumptions.
\end{itemize}

%% =====================================================================
%% ETHICS STATEMENTS -- keep only the statements relevant to your work.
%% =====================================================================
\section*{Ethics statements}

This study involved only the passive recording and analysis of environmental
audio from free-ranging animals in their natural habitat. It included no
animal experiments and no capture, handling, or manipulation of animals; no
human subjects; and no data collected from social-media platforms. [If
applicable, add: field recordings were collected under permit [number] issued
by [authority], in accordance with the relevant institutional and national
guidelines for non-invasive field research.]

%% =====================================================================
%% CREDIT AUTHOR STATEMENT
%% =====================================================================
\section*{CRediT author statement}

\textbf{Moshi Fu:} Conceptualization, Methodology, Software, Validation,
Formal analysis, Data curation, Writing -- original draft, Visualization.
\textbf{[Co-author Name]:} [Supervision, Resources, Writing -- review \&
editing, Funding acquisition].

%% =====================================================================
\section*{Acknowledgments}

[Acknowledge field teams, station maintainers, and anyone who contributed
reference annotations but does not meet authorship criteria.] This research did
not receive any specific grant from funding agencies in the public,
commercial, or not-for-profit sectors.

%% =====================================================================
%% DECLARATION OF INTERESTS -- keep BOTH statements; tick the correct box.
%% =====================================================================
\section*{Declaration of interests}

\noindent$\boxtimes$~The authors declare that they have no known competing
financial interests or personal relationships that could have appeared to
influence the work reported in this paper.

\medskip
\noindent$\square$~The authors declare the following financial
interests/personal relationships which may be considered as potential
competing interests:

%% =====================================================================
\section*{Supplementary material and/or additional information [OPTIONAL]}

[Optional. List any supplementary files here, e.g.\ example reference clips,
the \texttt{config.py} used, or sample detection CSVs. Delete this section if
not used.]

%% =====================================================================
\begin{thebibliography}{99}

\bibitem{cauzinille2024}
J. Cauzinille, B. Favre, R. Marxer, A. Rey, Applying machine learning to
primate bioacoustics: review and perspectives, Am. J. Primatol. (2024) e23666.
\url{https://doi.org/10.1002/ajp.23666}

\bibitem{sun2022}
Y. Sun, T.M. Maeda, C. Sol\'is-Lemus, D. Pimentel-Alarc\'on, Z. Bu\v{r}ivalov\'a,
Classification of animal sounds in a hyperdiverse rainforest using convolutional
neural networks with data augmentation, Ecol. Indic. 145 (2022) 109621.
\url{https://doi.org/10.1016/j.ecolind.2022.109621}

\bibitem{simonyan2015}
K. Simonyan, A. Zisserman, Very deep convolutional networks for large-scale
image recognition, in: International Conference on Learning Representations
(ICLR), 2015. \url{https://doi.org/10.48550/arXiv.1409.1556}

\bibitem{deng2009}
J. Deng, W. Dong, R. Socher, L.-J. Li, K. Li, L. Fei-Fei, ImageNet: a
large-scale hierarchical image database, in: IEEE Conference on Computer
Vision and Pattern Recognition (CVPR), 2009, pp. 248--255.
\url{https://doi.org/10.1109/CVPR.2009.5206848}

\bibitem{gemmeke2017}
J.F. Gemmeke, D.P.W. Ellis, D. Freedman, A. Jansen, W. Lawrence, R.C. Moore,
M. Plakal, M. Ritter, Audio Set: an ontology and human-labeled dataset for
audio events, in: IEEE International Conference on Acoustics, Speech and Signal
Processing (ICASSP), 2017, pp. 776--780.
\url{https://doi.org/10.1109/ICASSP.2017.7952261}

\bibitem{hershey2017}
S. Hershey, S. Chaudhuri, D.P.W. Ellis, J.F. Gemmeke, A. Jansen, R.C. Moore,
M. Plakal, D. Platt, R.A. Saurous, B. Seybold, M. Slaney, R.J. Weiss,
K. Wilson, CNN architectures for large-scale audio classification, in: IEEE
International Conference on Acoustics, Speech and Signal Processing (ICASSP),
2017, pp. 131--135. \url{https://doi.org/10.1109/ICASSP.2017.7952132}

\bibitem{lee2018}
K. Lee, K. Lee, H. Lee, J. Shin, A simple unified framework for detecting
out-of-distribution samples and adversarial attacks, in: Advances in Neural
Information Processing Systems (NeurIPS) 31, 2018, pp. 7167--7177.
\url{https://doi.org/10.48550/arXiv.1807.03888}

\bibitem{mcfee2015}
B. McFee, C. Raffel, D. Liang, D.P.W. Ellis, M. McVicar, E. Battenberg,
O. Nieto, librosa: audio and music signal analysis in Python, in: Proceedings
of the 14th Python in Science Conference (SciPy), 2015, pp. 18--25.
\url{https://doi.org/10.25080/Majora-7b98e3ed-003}

\end{thebibliography}

\end{document}
